Accurate binding of calcium to phospholipid bilayers by effective inclusion of electronic polarization

### 2017-10-18 Manuscript/manuscript.tex

\subsection{Electronic continuum correction for lipid bilayers}\label{section:ecc}
The lack of electronic polarizability in the standard MD simulation
force fields has been considered a higly relevant issue since the early days of
lipid bilayer simulations.
%Some pioneering simulation studies scaled
%the partial charges of atoms by one half to effectively include 
%polarizability \cite{jonsson86,egberts94}. Also, approaches that
%explicitly include polarizability has been introduced, but
%this has turned out to be practically complicated \cite{chowdhary13}.
In this work we circumvent the rather demanding explicit inclusion
of electronic polarization effects \cite{chowdhary13} by implicitly including
electronic polarizability in lipid
bilayer simulations by using the electronic continuum correction
(ECC)~\cite{leontyev11}. 
Technically, it is a similar approach to the phenomenological charge-scaling as applied in 
the early studies where a scaling factor one half was used \cite{jonsson86,egberts94}.
\todo{We should also cite papers where empirical scaling was used ionic liquids - but there the factor is not 0.5.}
However, the present concept of ECC is 
physically well justified and rigorously derived~\cite{leontyev09, leontyev10, leontyev11}.
%but significantly more simple than the explicit inclusion of electronic
%polarizability \cite{chowdhary13} both for derivation and application.

According to ECC, electronic
polarizability can be implicitly included in classical MD simulations by
placing all particles into a homogeneous dielectric continuum 
with a dielectric constant $\epsilon _{el}$, 
which is the electronic part of the dielectric constant of 
the media \cite{leontyev11}. Measurements of high frequency 
dielectric constant gives values of approximately $\epsilon _{el} \approx 2$ 
for almost any biomaterial \cite{some_original_work, leontyev11}.
Such a dielectric continuum can be easily included in standard MD simulation by
a formal transformation of partial charges 
\begin{equation}
  Q^{ECC} = f_q \cdot Q
\end{equation}
with a constant scaling factor $f_q = \epsilon _{el} ^{-1/2}$ 
effectively representing the newly introduced electronic continuum. 
Assuming globally a  high frequency dielectric constant as measured in water (corresponding to the square of the refraction index),  $\epsilon _{el} = 1.78$, results 
in a scaling factor of $f_q = 0.75$ \cite{some_orig_source, leontyev11}. This scaling factor has been
successfully used to improve the performance of force field for ions in solution \cite{kohagen14,kohagen16,??} which then agree quantitatively with neutron scattering data \cite{kohagen14,kohagen16, Pluharova2014}.

While the scaling factor of $f_q = 0.75$ for ions in water improves their performance and is physically justifiable within the ECC theory \cite{Leontyev2011},
it is not clear whether the same factor should be used for partial charges in molecules,
e.g., lipids in our case. Unlike the total charge of an atom or molecules, atomic partial charges within each molecule are not physical observables.
There are several schemes for the assignment of partial charges for biomolecules. \cite{Hu2007}
Currently, the most commonly employed scheme is the restrained electrostatic potential (RESP) method~\cite{RESP_paper,Singh1984}.
%
%PAVEL: i do not understand what you mean by "ensemble" - resp can be done for a single configuration of a single molecule.
%In any case, tehre is no need to explain here how RESP works so i would drop this part (including the next two sentences).
%
%SAMULI: I commented out the suggested sentences
%which aims to reproduce the electrostatic potential obtained from quantum mechanical calculations around an ensemble the target molecules. 
%The fixed partial atomic charges obtained by this method are hence 
%the best fit of pint charges reproducing the electrostatic potential around provided target conformations. 
%Such a scheme is well suited for the application of ECC, 
%which targets at improving electrostatic interactions using a mean field correction. 
In practice, partial charges currently implemented in force fields may already 
include  to some extend some of the solvent electronic polarizability effects, i.e.,
the RESP charges are often scaled to fit some experimental observables \todo{This needs a citation}. 
Thus, we expect that the application of the ECC scaling factor, $f_q$, to the molecular partial charges
included in the available force fields does not necessarily have to follow the relation $f_q = \epsilon _{el} ^{-1/2}$, 
but instead it lies between 0.75 (no electronic polarizability in the partial charge calculation)
and $1$ (full electronic polarizability already included in the partial charge calculation).
%
% This is not really important for the paper:
%
%However, the RESP method has several remedies;  %and these are specific to RESP
%RESP charges show a dependence on the orientation and position
%and it does not provide numerically stable partial charges of burried atoms;~\cite{Hu2007}
%the RESP-derived values depend, for example, on the conformational
%sampling of the molecules and description of solvent 
%in the calculations~\cite{RESP_paper, Singh1984, Hu2007, maciejewski14}. 
%
%Hirshfeld-E partitioning: \cite{Verstraelen2013}
%In cases where the partial charges have some physical meaning 
%(e.g. approximate electrostatic potential as in RESP~~\cite{RESP_paper,Singh1984}),
%
%

In this work, we develop a phospholipid model for classical MD simulations that
accurately describes the lipid head group response to varying concentrations
of monovalent and divalent cations. This is a biologically highly relevant membrane feature, which is
poorly reproduced by currently available models which can affect not only on membrane properties
in the presence of ions but also modulates the interaction with charged moieties in the surface.
Importantly, this response from simulations can be accurately
compared against experimental NMR data \cite{akutsu81,altenbach84,scherer89},
as discussed in Ref. \citenum{catte16} in section \ref{section:electrometer}.
To this end, we empirically explore the scaling factor parameter space, $f_q \in [0.75, 1.0]$ for the Lipid14~\cite{dickson14} force field.
We selected this force field as a starting point because its response to bound ions was apparently the most realistic against NMR data
in recent work by NMRlipids project (see Fig.~5 in Ref.~\citenum{catte16}).
Also glycerol backbone and head group structures in Lipid14 model were
relatively realistic when compared with other state of the art lipid models \cite{botan15}.
The ECC correction was applied to Lipid14 parameters by scaling partial charges of the head group,
glycerol backbone and carbonyl regions, which are the most polar parts in lipids and
are expected to have the largest contribution to the cation binding
However, we do not modify the hydrocarbon chain parameters, because they
do not come in contact with salt ions and are
already highly optimized and give generally a good description for
hydrophobic part of lipid bilayers in most lipids models, including Lipid14 \cite{ollila16}.
This is in contrasts with the behavior in glycerol backbone and head group order
parameters which call for improvements in all available lipid models~\cite{botan15}.

Exploring different scaling factor values, we found out that ion binding and
related head group order parameter responses become weakened in general. 
The optimal behaviour of ion binding was observed with
the scaling factor of $f_q = 0.8$. % is optimal for applying ECC on Lipid14 model. 
\todo{JOE: following discussion shall be modified in the enlightenment of our recent ECC-discussions.}
Interestingly, this scaling factor is in line with the estimate
given by ``implicitly polarized charges'' (IPolQ) \cite{ipolq2013}
combined with RESP calculations in vacuum and implicit solvent reported
in \cite{maciejewski14}. IPolQ charges are obtained as the average of
partial charges given by RESP calculation~\cite{RESP_paper}
in vacuum and in a solvent. Applying the scaling factor of 0.75 to
IPolQ charges calculated from the data in Ref.~\cite{maciejewski14},
gives similar partial charges to ones obtained by scaling Lipid14 charges
with a factor 0.8.
%-----\footnote----------
  %{Depending on which QM method you arrive at values from 0.76 to 0.83, averages across atom
  %types being around 0.78--0.80. Even though the methods are almost identical, authors of Lipid 14 find lower
  %partial charges in vacuum than here -- so I prefer the higher value. As the choice of charges is arbitrary anyway,
  %I use 0.80 as an approximate round value, which accounts for the already lower charges in Lipid14. 
  %The use of 0.78---.79 might be more appropriate, though.}

While, the charge scaling improved the behaviour of lipid-ion interactions,
it reduced the area per molecule of the lipid bilayer without ions below experimental
values. Simulations with Lipid14 parameters having partial charges of head group, glycerol
backbone and carbonyls scaled with 0.8 gave the area per molecule value of $\approx$60~\AA$^2$,
which is smaller than the experimental value 64.3~\AA$^2$ (\cite{}\todo{missing REF for APL experiment})
and the original Lipid14 value $(65.6 \pm 0.5)$~\AA$^2$~\cite{dickson14}.
The decrease of the area per lipid was found to arise from a lower hydration of the lipid head group region,
which can be explained by the increased solvation free energy due to the lower polarity
of molecules with scaled charges. The hydration can be increased back
by reducing the effective radius of atoms by changing the 
the $\sigma$ parameters in the Lennard-Jones potential for the selected atoms
similarly as done for free ions in solution \cite{kohagen14,kohagen16,Pluharova2014}.
\todo{We should discuss how this can potentially affect the intermolecular interaction when mixing scaled and non scaled molecules. \\
  JOE: I think that we rather increasingly see that there's nothing like ``fully non-scaled'' with the exception of ions with integer
  charges. So the discussion shall be rather more about the interaction of our ``scaled'' (I'd still rather call it ECC-corrected or whatever)
  and ``semi-scaled'' models. \\
SAMULI: There is now a paragraph in the conclusions, which mentions this topic.}
This decreases the solvation free energy by allowing water molecules to
approach closer to lipid atoms and have stronger electrostatic interactions with them.
After reducing the $\sigma$ parameters by a factor of $f_\sigma = 0.89$
for the same atoms for which charges were scaled, the area per molecule value was
back in agreement with the experimental value (see Table~\ref{tab:apls}).

>>>>

\subsection{Electronic continuum correction for lipid bilayers}\label{section:ecc}
The lack of electronic polarizability in the standard MD simulation
force fields has been considered a higly relevant issue since the early days of
lipid bilayer simulations.
%Some pioneering simulation studies scaled
%the partial charges of atoms by one half to effectively include 
%polarizability \cite{jonsson86,egberts94}. Also, approaches that
%explicitly include polarizability has been introduced, but
%this has turned out to be practically complicated \cite{chowdhary13}.
In this work we circumvent the rather demanding explicit inclusion
of electronic polarization effects \cite{chowdhary13} by implicitly including
electronic polarizability in lipid
bilayer simulations by using the electronic continuum correction
(ECC)~\cite{leontyev11}. 
Technically, it is a similar approach to the phenomenological charge-scaling as applied in 
earlier studies \cite{jonsson86,egberts94}.
\todo{We should also cite papers where empirical scaling was used ionic liquids DOI: 10.1002/anie.201308760}
The present concept of ECC is, however, 
physically well justified and rigorously derived~\cite{leontyev09, leontyev10, leontyev11}.
%but significantly more simple than the explicit inclusion of electronic
%polarizability \cite{chowdhary13} both for derivation and application.

Following the ECC concept, electronic
polarizability can be effectively included in classical MD simulations by
placing all particles into a homogeneous dielectric continuum 
with a dielectric constant $\epsilon _{el}$, 
which is the electronic part of the dielectric constant of 
the media \cite{leontyev11}. Measurements of high frequency 
dielectric constant gives values around 2 
for almost any biologically relevant material \cite{some_original_work, leontyev11}.
Following the Coulomb's law, such a dielectric continuum can be easily included in standard MD simulation by
a scaling of charges 
\begin{equation}
  Q^{ECC} = f_q \cdot Q
\end{equation}
with a constant scaling factor $f_q = \epsilon _{el} ^{-1/2}$ 
effectively representing the newly introduced electronic continuum. 
Given the  high frequency dielectric constant as measured in water (corresponding to the square of the refraction index) of  $\epsilon _{el} = 1.78$,  
the scaling factor for aqueous ions is $f_q = 0.75$ \cite{some_orig_source, leontyev11}. This scaling factor has been
successfully used to improve the performance of force field for ions in solution \cite{kohagen14,kohagen16,??} which then agree quantitatively with neutron scattering data \cite{kohagen14,kohagen16, Pluharova2014}.

While the scaling factor of $f_q = 0.75$ for ions in water improves their description and is physically well justified within the ECC theory \cite{Leontyev2011},
it is nota priori  clear whether the same factor should be used for partial charges in molecules,
e.g., lipids in our case. Unlike the total charge of an atom or molecules, atomic partial charges within molecules are not physical observables.
There are thus several schemes for the assignment of partial charges for biomolecules. \cite{Hu2007}
Currently, the most commonly employed scheme is the restrained electrostatic potential (RESP) method~\cite{RESP_paper,Singh1984}.
%
%PAVEL: i do not understand what you mean by "ensemble" - resp can be done for a single configuration of a single molecule.
%In any case, tehre is no need to explain here how RESP works so i would drop this part (including the next two sentences).
%
%SAMULI: I commented out the suggested sentences
%which aims to reproduce the electrostatic potential obtained from quantum mechanical calculations around an ensemble the target molecules. 
%The fixed partial atomic charges obtained by this method are hence 
%the best fit of pint charges reproducing the electrostatic potential around provided target conformations. 
%Such a scheme is well suited for the application of ECC, 
%which targets at improving electrostatic interactions using a mean field correction. 
By construction, partial charges currently implemented in force fields may already 
include  to some extend the solvent electronic polarizability effects, i.e.,
the RESP charges are often scaled to fit certain experimental observables \todo{This needs a citation - Joe please pick 1-2 papers}. 
Thus, a consistent application of the ECC scaling factor, $f_q$, to the molecular partial charges
included in the available force fields does not necessarily have to follow the above relation $f_q = \epsilon _{el} ^{-1/2}$, 
but instead it lies between 0.75 i.e., (no electronic polarizability included in the original partial charges)
and $1$ (i.e., electronic polarizability fully included in the original partial charges).
%
% This is not really important for the paper:
%
%However, the RESP method has several remedies;  %and these are specific to RESP
%RESP charges show a dependence on the orientation and position
%and it does not provide numerically stable partial charges of burried atoms;~\cite{Hu2007}
%the RESP-derived values depend, for example, on the conformational
%sampling of the molecules and description of solvent 
%in the calculations~\cite{RESP_paper, Singh1984, Hu2007, maciejewski14}. 
%
%Hirshfeld-E partitioning: \cite{Verstraelen2013}
%In cases where the partial charges have some physical meaning 
%(e.g. approximate electrostatic potential as in RESP~~\cite{RESP_paper,Singh1984}),
%
%

In this work, we develop a phospholipid model for classical MD simulations that
accurately describes the lipid head group response to varying concentrations
of monovalent and divalent cations. This is a biologically highly relevant membrane feature, which is
poorly reproduced by currently available models. This can affect not only on membrane properties
in the presence of ions but also modulate interactions with charged moieties at the interface.

Importantly, results from simulations can be directly
compared against experimental NMR data \cite{akutsu81,altenbach84,scherer89},
as discussed in detail in Ref. \citenum{catte16}.
To this end, we explore the above discussed scaling factor parameter space, $f_q \in (0.75, 1.0)$ for the Lipid14~\cite{dickson14} force field.
We selected this force field as a starting point because its response to bound ions was apparently the most realistic against NMR data
in a recent test within NMRlipids project (see Fig.~5 in Ref.~\citenum{catte16}).
Moreover, the glycerol backbone and head group structures in Lipid14 model were
more realistically described compared with other state of the art lipid models \cite{botan15}.
The ECC correction was applied to Lipid14 parameters by scaling partial charges of the head group,
glycerol backbone, and carbonyl regions, which are the most polar parts in the  lipids and
are thus expected to have the largest contribution to the cation binding
In contrast, we do not modify the hydrocarbon chain parameters, as they
do not come in direct contact with salt ions and are
already highly optimized to provide a good description for the
hydrophobic part of lipid bilayers\cite{ollila16}.
This is in contrasts with the behavior in glycerol backbone and head group order
parameters in the presence of ions, which call for improvements in all available lipid models~\cite{botan15}.

Exploring different values of the scaling factor, we found out that ion bindings and the
related head group order parameter responses become weakened in general. 
The optimal behaviour  was observed for a
scaling factor of $f_q = 0.8$.This value is only slighly higher than the original scaling factor of 0.75, which implies
that electronic polarization effects are only to a small extent effectively accounted for in the Lipid14 force field. 
%-----\footnote----------
  %{Depending on which QM method you arrive at values from 0.76 to 0.83, averages across atom
  %types being around 0.78--0.80. Even though the methods are almost identical, authors of Lipid 14 find lower
  %partial charges in vacuum than here -- so I prefer the higher value. As the choice of charges is arbitrary anyway,
  %I use 0.80 as an approximate round value, which accounts for the already lower charges in Lipid14. 
  %The use of 0.78---.79 might be more appropriate, though.}

While, the above charge scaling improved the description of lipid-ion interactions,
it also led to a reduction of the area per molecule of the lipid bilayer in water below experimental
values. Simulations with Lipid14 parameters having partial charges of head group, glycerol
backbone and carbonyls scaled with 0.8 gave the area per molecule value of $\approx$60~\AA$^2$, as well as
which is somewhat smaller than the experimental value of 64.3~\AA$^2$ (\cite{}\todo{missing REF for APL experiment})
the Lipid14 value of $(65.6 \pm 0.5)$~\AA$^2$~\cite{dickson14}.
The decrease of the area per lipid is found to arise from a reduced hydration of the lipid head group region,
due to the lower polarity of molecules with scaled charges. This can be compensated
by reducing the effective radius of atoms with scaled charges by changing the 
the $\sigma$ parameters in the Lennard-Jones potential, as done previously for ions in solution \cite{kohagen14,kohagen16,Pluharova2014}.
\todo{We should discuss how this can potentially affect the intermolecular interaction when mixing scaled and non scaled molecules. \\
  JOE: I think that we rather increasingly see that there's nothing like ``fully non-scaled'' with the exception of ions with integer
  charges. So the discussion shall be rather more about the interaction of our ``scaled'' (I'd still rather call it ECC-corrected or whatever)
  and ``semi-scaled'' models. \\
SAMULI: There is now a paragraph in the conclusions, which mentions this topic.}
After reducing these $\sigma$ parameters by a factor of $f_\sigma = 0.89$, the area per molecule is restored
back close to the experimental value (see Table~\ref{tab:apls}).
